---
title: 'Simon''s Fish Gut Content: SAP Evaluation'
author: "Evan Krell"
date: "August 10, 2017"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
suppressPackageStartupMessages(library("Vennerable"))
library (knitr)
library (ggplot2)
library(MASS)

# Load data 
sapTbl <- read.csv(file = "classification_exp2.csv", header = TRUE)
fastaLen = 199

# Delete unneeded rows
sapTbl$BranchCount <- NULL # Is not needed since equivalent to 'SpeciesCount'
sapTbl$TreeDepth <- NULL # Completetly dependent on the assignment rank level. Not interesting?
# Type conversions
sapTbl$SpeciesLevel <- as.logical (sapTbl$SpeciesLevel)
sapTbl$HasHomologuesSuspiciousRecords <- as.logical(sapTbl$HasHomologuesSuspiciousRecords)
sapTbl$DatabaseNotExhausted <- as.logical (sapTbl$DatabaseNotExhausted)
```

[Statistical Assignment Package (SAP)](https://github.com/kaspermunch/sap) was used for taxonomic assignment of the gut contents of several Gulf of Mexico species. 
These samples were collected by [Jim Simons' lab](http://gomexsi.tamucc.edu/gomexsi-partners/dr-james-simon/) and processed by the [Genomics Core Lab](http://genomics.tamucc.edu/). 

The companion script evaluateSAP/evaluateSAP.py has been used to quantify several attributes related to the SAP results. 
The goal is to analyze this table to learn which assignments are plausible and what factors affect the ability to make a species-level assignment. 

SAP was run with the default parameters. The most important parameters are these:

| Parameter       | Selected | Description                         | 
| --------------- | -------- | ----------------------------------- |
| Assignment | Barcoder | A Bayesian method for taxonomic assignment that used Markov chain monte carlo. |
| Posterior Probability Cutoff | 0.95 | Only report assignments with posterior probability at least 0.95 |
| Min Significance | 10 | "Lower E-value bound for accepting homologue sequences" |
| Significance | 0.10 | "E-value cutoff for significant hits." |
| **Min Identity** | 0.97 | "Minimum global alignment similarity of best blast hit." |
| Best Hits | 30 | "Maximal number of homologues to add to alignment purely based on significant E-value." |

Command used:

      sap --project Exp2 --database ../nr_mito.SAP2.fasta \
          --minidentity 0.97 \
          ../SAPexplore-OTU.fasta


## Data overview
```{r overview, fig.width=10,fig.height=5, echo = FALSE}

# Display first 10 rows (the 'head') as example
sapTblHead <- head(sapTbl)
kable (sapTblHead)

# Summarize data set
summary(sapTbl[2:ncol(sapTbl),])
```

### Species Assignment
```{r speciesAssignment, fig.width=10,fig.height=5, echo = FALSE}
numSpeciesLevel = nrow (sapTbl[sapTbl$SpeciesLevel  == TRUE, ])
numHigherLevel = nrow (sapTbl[sapTbl$SpeciesLevel  == FALSE, ])
numUnassigned = fastaLen - (numSpeciesLevel + numHigherLevel)
slices <- c(numSpeciesLevel, numHigherLevel, numUnassigned)
labels <- c("Species-Level", "Nonspecies-Level", "Unassigned")
pie(slices, labels = labels, main="Assignment rank")

```

### Compare posterior probabilties
```{r comparePosteriorProbabilities, fig.width=10,fig.height=5, echo = FALSE}
# Box plots comparing whether or not species rank
VariableNames <- c ("Species Assigned", "Higher Rank Assigned")
# Compare posterior probabilities 
mm_probabilities = melt(sapTbl, id.vars = 'SpeciesLevel', measure.vars = c ('SpeciesProbMax', 'SpeciesProbMin'))
ggplot(mm_probabilities)+geom_boxplot(aes(x=SpeciesLevel, y=value, color=variable))
```

### Compare tree topology
```{r compareTreeTopology, fig.width=10,fig.height=5, echo = FALSE}
# Compare tree topology
mm_branches = melt(sapTbl, id.vars = 'SpeciesLevel', measure.vars = c ('BranchLeafCount', 'BranchInteriorCount'))
ggplot(mm_branches)+geom_boxplot(aes(x=SpeciesLevel, y=value, color=variable))
```

### Compare homologues
```{r compareHomologues, fig.width=10,fig.height=5, echo = FALSE}
# Compare homologues
mm_homologues = melt(sapTbl, id.vars = 'SpeciesLevel', measure.vars = c ('HomologueCount','HomologuesSuspiciousRecordsCount'))
ggplot(mm_homologues)+geom_boxplot(aes(x=SpeciesLevel, y=value, color=variable))

```

### Database not Exhausted 
```{r databaseNotExhausted, echo = FALSE}
numDBnotExhausted = nrow (sapTbl[sapTbl$DatabaseNotExhausted == TRUE, ]) 
numDBnotExhausted_SpeciesLevel = length (which(sapTbl[sapTbl$DatabaseNotExhausted == TRUE, ]$SpeciesLevel == TRUE))
numDBnotExhausted_HigherLevel = length (which(sapTbl[sapTbl$DatabaseNotExhausted == TRUE, ]$SpeciesLevel == FALSE))
slices <- c(numDBnotExhausted_SpeciesLevel, numDBnotExhausted_HigherLevel)
labels <- c("Species-Level", "Nonspecies-Level")
pie(slices, labels = labels, main="Assignment rank when database not exhausted")

numDBnotExhausted_Suspicious = length(which(sapTbl[sapTbl$DatabaseNotExhausted == TRUE, ]$HasHomologuesSuspiciousRecords == TRUE))
numDBnotExhausted_NotSuspicious = length(which(sapTbl[sapTbl$DatabaseNotExhausted == TRUE, ]$HasHomologuesSuspiciousRecords == FALSE))
slices <- c(numDBnotExhausted_Suspicious, numDBnotExhausted_NotSuspicious )
labels <- c("Suspicious", "Not Suspicious")
pie(slices, labels = labels, main="Whether or not suspicious homologues\n are present when database is not exhausted\n(But assignment at some rank made)")

```



